---
title: "Titanic to Classification"
author: "Author"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: no
    toc_depth: '6'
  pdf_document:
    toc: no
    toc_depth: '6'
  html_notebook:
    fig_caption: no
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
    toc: no
    toc_depth: '6'
urlcolor: blue
subtitle: ""
fontsize: 10pt
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=FALSE, message=FALSE)
```

# Overview

On 15th April, 1912 the Titanic Sank on its return voyage to New York. There were an estimated 2224
passengers and crew on board, with over 1500 fatalities from the sinking.\
The Titanic dataset provides a sample of passengers, with some basic details: their names, ages, gender, class, and several others. We are predicting a binary response: whether the passenger survived or died.

# Set-up

## Load the titanic and titanic_hidden datasets

```{r}
titanic <- read.csv("titanic.csv", row.names=1)
titanic_hidden <- read.csv("titanic.csv", row.names=1)[1:200,]
```

# Lab Steps

## Clean the dataset for use in R

```{r}
library(caret)
library(tidyverse)
library(rsample)
```


```{r}
titanic <- titanic %>% select(-Name)
titanic_hidden <- titanic_hidden %>% select(-Name)
titanic$Survived <- factor(titanic$Survived, labels = c("Deceased", "Survived"))
titanic_hidden$Survived <- factor(titanic_hidden$Survived, 
                                  labels = c("Deceased", "Survived"))
```

## Display the summary from a logistic regression: response Survived and include every other variable as predictors

```{r}
# define training control
train_control <- trainControl(method = "cv", number = 10)
# train the model on training set
logistic_reg_model <- train(Survived ~ .,
                   data = titanic,
                   trControl = train_control,
                   method = "glm",
                   family=binomial())
```


```{r}
logistic_reg_model_summary <- summary(logistic_reg_model)
logistic_reg_model_summary
```

## Interpret the meaning of the Male gender indicator variable coefficient

```{r}
# Coefficients
logistic_reg_model_summary$coefficients[3,,drop =F]
```

The `Sexmale` coefficient is statistically significant (associated with a p-value < 0.05), for which the logistic regression model suggests that passenger being of male sex does in fact influence the survival chances, and as it is a negative number, implies that passenger being of male sex decreases the survival chance.

```{r}
coeff_logistic_reg_model <- coef(logistic_reg_model$finalModel)
```

When interpreting the differences across genders, it is much easier to infer from $\text{exp}(\beta_{\text{Sexmale}})$ = `r exp(coeff_logistic_reg_model['Sexmale'])` that represents the odds ratio associating passenger gender to rate survival.

A passenger being of the male gender has an associated `r 100*(1-exp(coeff_logistic_reg_model['Sexmale']))` % (1 -$\text{exp}(\beta_{\text{Sexmale}})$ = `r 1-exp(coeff_logistic_reg_model['Sexmale'])` ) less chance of survival compared to female gender passengers.

## Use your model to predict the status of the passengers in the titanic_hidden dataset

```{r}
log_prediction <- predict(logistic_reg_model, titanic_hidden)
confusionMatrix(log_prediction, titanic_hidden$Survived)
```

## Now use caret to set up a KNN with response Survived and all other variables, then predict the status of passengers in the titanic_hidden dataset

```{r}
set.seed(400)
knn_reg_model <- train(Survived ~ ., 
                data = titanic, 
                method = "knn", 
                trControl = train_control, 
                preProcess = c("center","scale"), tuneLength = 20)
```


```{r}
#Output of kNN fit
knn_prediction <- predict(knn_reg_model, titanic_hidden)
confusionMatrix(knn_prediction, titanic_hidden$Survived)
```

## Which model performs best for "Accuracy"?

```{r}
accuracy_data_frame = data.frame(
  Model = c('Logistic Regression', 'KNN'),
  Accuracy = c(
    confusionMatrix(log_prediction, titanic_hidden$Survived)$overall['Accuracy'],
    confusionMatrix(knn_prediction, titanic_hidden$Survived)$overall['Accuracy']
  )
)
accuracy_data_frame[order(-accuracy_data_frame$Accuracy),]
accuracy_data_frame
```

`r paste0('Model that performs best for Accuracy: ', accuracy_data_frame[1,1])`